---
title: "ISTA 320 Final Project"
author: "A. Jess White"
date: "5/6, Spring 2022"
output: html_document
---

# Data Reading and Wrangling

```{r setup, include=FALSE}
# Load libraries
library(ggplot2)
library(dplyr)
library(maps)

# Read in raw data
raw_cars_data <- read.csv("data/cars.csv")

# Slight bit of cleanup: This ID is associated with the reward pool, which are cars we have not seen and must get rid of.
clean_cars_data <- raw_cars_data[!(raw_cars_data$ID == 942407134514135050),]

# Some cars are marked as unknown year or as a range, like "197x".
# Let's turn these into NAs
clean_cars_data$Year <- as.numeric(crude_cars_data$Year)

# We want to analyze only daily driving vehicles. Let's filter out some odd brands
# We don't want buses, semitrucks, or work vehicles.
clean_cars_data <- clean_cars_data[!(clean_cars_data$Make == "Blue Bird" |
                                     clean_cars_data$Make == "Freightliner" |
                                     clean_cars_data$Make == "Gillig" |
                                     clean_cars_data$Make == "International" |
                                     clean_cars_data$Make == "Kenworth" |
                                     clean_cars_data$Make == "Mitsubishi Fuso" |
                                     clean_cars_data$Make == "Motor Coach Industries" |
                                     clean_cars_data$Make == "Peterbilt" |
                                     clean_cars_data$Make == "Workhorse Custom Chassis"),]

# Finally, we want to make sure brand names are standardized.
clean_cars_data <- clean_cars_data %>%
  mutate(Make = case_when(Make == "KIA" ~ "Kia",
                          Make == "RAM" ~ "Ram",
                          TRUE ~ Make))

```

# Data Set Description

This is in the README.md!

# Data Questions

Timeseries Scatter Plot
1. What is the average age of the cars that are on the road today? What are the newest and oldest cars still driving?

Bar graph
2. Which car manufacturer is the most popular here in Tucson? Which manufacturer is the least popular?

World map
3. Where do these cars originate from? What is the most popular country of origin? In general, where are the local cars coming from and in what distribution?


# Timeseries Scatter Plot

Build a line plot to answer question #1.

```{r}

# Remove NA years
year_corrected_data <- clean_cars_data[!(is.na(clean_cars_data$Year)),]

# Make a table with years and their counts
year_counts <- data.frame(table(year_corrected_data$Year))

# Rename the variables
year_counts = rename(year_counts, Year = Var1, count = Freq)

# Turn the fct into a dbl so we can scale the graph
year_counts$Year <- as.numeric(as.character(year_counts$Year))

# Check to see that it worked
glimpse(year_counts)

# Create a timeseries scatter plor
year_counts %>%
  ggplot(aes(x = Year,
             y = count)) +
  geom_point() +
  geom_label(data = filter(year_counts, count == max(count)), aes(label = count)) +
  scale_x_continuous()
```

As we can see in this scatter plot, the average age of cars is around 5 years old. In our dataset,
118 cars have been spotted that were made in 2017, and the years around it are also in the hundreds.

The newest cars are from 2022, and they're somewhat uncommon as there have been about 40 spotted.

The oldest car seen on the road is from 1940: an 82 year old car! This particular car is a 1940 Ford Coupe.

# Bar Plot

Build a bar plot to answer question #2.

```{r}

# Make a dataframe with the car makes and their counts
make_counts <- data.frame(table(clean_cars_data$Make))

# Rename the auto-assigned variables
make_counts <- rename(make_counts, Make = Var1, Count = Freq)

# Make a plot that makes manufacturers easier to find
make_counts %>%
  ggplot(aes(x = Make,
             y = Count)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Number of cars by manufacturer: Alphabetical order")

# Make another plot, but with clearer comparisons between counts
make_counts %>%
  ggplot(aes(x = reorder(Make, Count),
             y = Count)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Number of cars by manufacturer: Ordered by count")

```

It is clear by both of these charts that the most popular brand is Chevrolet. It is close though, with Ford and Toyota competing at the same level. The remaining brands quickly drop in popularity.

The least popular brands are Alfa Romeo, Bentley, Datsun, and Plymouth, at least out of the ones we were able to find. There are car manufacturers that we have yet to see, but these are the least popular of the ones we have found. This makes sense too: Alfa Romeos are foreign and expensive cars, Bentleys are incredibly expensive and luxurious, and Plymouth and Datsun have both been out of business for a very long time.

# World Map

Build a world map to answer question #3.

```{r}
# Build map template
world <- map_data("world")

# View which entries do not match
clean_cars_data %>%
  filter(!(Birthplace %in% unique(world$region))) %>%
  pull(Birthplace)

# Manually correct some countries
clean_cars_data <- clean_cars_data %>%
  mutate(Birthplace = case_when(Birthplace == "United States" ~ "USA",
                                Birthplace == "UNITED STATES" ~ "USA",
                                Birthplace == "United Kingdom" ~ "UK",
                                Birthplace == "MEXICO" ~ "Mexico",
                                Birthplace == "JAPAN" ~ "Japan",
                                TRUE ~ Birthplace))

# Record counts to be displayed on the map
country_counts <- data.frame(table(clean_cars_data$Birthplace))
# Remove invalid values
country_counts <- country_counts[!(country_counts$Var1 == "" | country_counts$Var1 == "N/A"),]
# Combine datasets
country_counts_world <- world %>%
  left_join(country_counts,
            by = c("region" = "Var1"))


country_counts_world %>%
  ggplot(aes(x = long,
             y = lat,
             fill = Freq,
             map_id = region)) +
  geom_map(map = world) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  ggtitle("Number of local cars manufactured by country") +
  scale_fill_continuous(name = "Cars manufactured",
                        low = "cornsilk",
                        high = "salmon4")
```

As you can see by this graph, there isn't much variety when it comes to car manufacturers. There isn't much competition either, as the United States manufactures the most local cars by a landslide. Canada, Japan, Mexico, and South Korea all produce a sizable portion of the local cars, but not enough to compete with the USA. The remaining countries, like Australia, Sweden, the UK, and Turkey all have only a handful of cars exported to Tucson.

Overall, we can see that the domestic makes (e.g. Ford, Chevrolet, Dodge) and foreign makes that have factories in the USA (e.g. Toyota) dominate the car market here in Tucson.